A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
more

% change?
\newcommand{\thO}{T_{o}}
\newcommand{\thD}{T_{d}}

@@

\section{The Theory of (Co)datatypes}
\label{sec:the-theory-of-co-datatypes}

@@

%\paragraph{Characterization.}
Datatypes can be defined semantically as the initial model of the
selector--constructor equations. This is the view followed by Barrett et al.\
\cite{barrett-et-al-2010}.
\begin{report}
A drawback of this approach is that it does
not naturally account for selectors applied to wrong constructors. Barrett et
al.\ address this by parameterizing the construction by default values, but
this gives rise to spurious equalities between unrelated terms---e.g.,
$\const{s}_{11}(\const{C}_2) \teq \const{s}_{11}(\const{C}_3)$. This flaw
could be corrected, but the added complexity seems to suggest that selectors
are better characterized axiomatically.

@@

Datatypes are additionally characterized by an induction axiom schema\begin{report}:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\,i,j}\, \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\,k}\, \mathit{IH}_{i\negvthinspace j}^k[x_k]\bigr) \longrightarrow P_i[C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\,i}\, P_i[v_i]$}
\DP
\end{aligned}
\]
where the induction hypothesis $\mathit{IH}_{i\negvthinspace j}^k(x)$
denotes either $P_{\iPrime}(x)$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$\end{report}.
%
The axiom schema ensures that the interpretation of datatypes
is a standard model. For example,
for a datatype of natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction prohibits nonstandard models, which could contain cyclic values---e.g.,
an $n$ such that $n \teq \const{S}(n)$---or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.
%
\begin{report}\par\end{report}
%
For codatatypes, the dual notion is called coinduction. \begin{report}It depends on
witnesses $R_i$ that are required to be bisimulations:
%
\[
\begin{aligned}[t]
\text{Coinduction:}\kern.8em %%% TYPESETTING: should be \quad
\AXC{\strut$\begin{gathered}\textstyle \bigwedge_{\,i}\, R_i[v_i, w_i] \\[-\jot]\textstyle
\bigwedge_{\,i}\, \forall v\; w.\;\, R_i[v, w] \longrightarrow
  \bigwedge_{\,j}\, \const{d}_j(v) \teq \const{d}_j(w)
  \mathrel\land
  \const{d}_j(v) \longrightarrow \bigwedge_{\,k}\, \const{s}_{i\negvthinspace j}^{\,k}(v) \sim \const{s}_{i\negvthinspace j}^{\,k}(w)
\end{gathered}$}
\UIC{\strut$\bigwedge_{\,i}\, v_i \teq w_i$}
\DP
\end{aligned}
\]
where $x \sim y$ denotes either $R_{\iPrime}[x, y]$ if there exists some
$\iPrime$ such that the formula is type-correct or $x \teq y$ otherwise.
\end{report}%
This axiom schema guarantees that two values that yield the same
observations must be equal, where the observations are made by using the
selectors and discriminators.
%
\begin{report}\par\end{report}
  %
In addition, codatatypes are guaranteed to contain all values corresponding to
infinite ground constructor terms. In general, this cannot be captured by a
first-order axiomatization, since there may be uncountably many of them.
\begin{report}
For example, $\ty{stream}_{\,\ty{int}}$ is isomorphic to the uncountable
function space $\ty{nat} \to \ty{int}$.
\end{report}

@@

This section presents a decision procedure for \rem{introduce ``$\thD$" in previous section?}
determining the $\thD$-satisfiability of sets of constraints built from a fixed signature $\Sig$.
We will make the following assumptions about $\Sig$.
First, all codatatypes in $\Codata$ are corecursive.
This is not restrictive, since checking the $\thD$-satisfiability for constraints involving a non-corecursive co-datatype $\tau$ can be trivially reduced to
checking the satisfiability of the same constraint involving a (non-recursive) datatype of the same form as $\tau$. \rem{revise?}
Second, all types $\tau \in \Nondata$ have infinite cardinality.
This also is not restrictive, since each interpreted type $\tau \in \Nondata$ having finite cardinality $n$
can be replaced by a enumeration datatype with $n$ constructors
\footnote{Doing so simplifies the presentation of the calculus, but is not done in practice. \rem{More?}}.
Additionally, since $\Ec$ is ground, we know that $\Ec$ cannot entail an upper bound on the cardinality of any uninterpreted type $\tau \in \Nondata$, 
and thus without loss of completeness we may treat all other types as infinite.

Our calculus consists of derivation rules that operate on a set $\Ec$ of equalities and disequalities between $\Sig$-terms,
where $\Ec$ is interpreted as a conjunction.
We write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
are simplified to $t_i$, and moreover assume that all %additional
constraints added to $\Ec$ are normalized in this manner.

@@

A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains an occurrence of a bound variable.
The notation $t \aequiv u$, expressing $\alpha$-\emph{equivalence},
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables.
For example, 
%$\mu x. \const{C}( x ) \aequiv \mu y. \const{C}( y )$ and
$\mu x.\; \const{D}( y, x ) \aequiv \mu z.\; \const{D}( y, z )$,
but
$\mu x.\; \const{C}( x ) \not\aequiv \mu x.\; \const{D}( y, x ) \not\aequiv \mu x.\; \const{D}( z, x )
\not\aequiv \mu y.\; \const{D}( y, x )$.
Two $\mu$-terms may represent the same value while not being $\alpha$-equivalent, as in
$\mu x.\; \const{C}( \const{C}( x ) ) \not\aequiv \mu x.\; \const{C}( x )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

Intuitively, a $\mu$-term $t$ of type $\tau$ describes a class of values of $\tau$, 
%either in the case when $\tau$ is a co-datatype or datatype type,
where a cyclic $t$ describes an infeasible class of values when
%$\tau \not\in \Codata$.
$\tau \in \Data$.
At the beginning of this step, 
we compute a mapping $\Val$ from equivalence classes to $\mu$-terms 
that describe the class of values that terms in that equivalence class can take in the models of $\Ec$. 
For each equivalence class $\ec{x}$, we associate a fresh variable $v_{\ec{x}}$ of the same type as $x$ not occurring in $\tEc$,
and initially say that $\Val$ contains $\ec{x} \mapsto v_{\ec{x}}$,
in other words, there are initially no constraints on the values for any equivalence class $\ec{x}$.
Then, $\Val$ is modified by applying the following update exhaustively:
\[
\inferrule{
  v_{\ec{x}} \in FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} ) \}
}
\]
Here, we write $FV( \Val )$ to denote the \emph{free variables} occurring in $\Val$, which
excludes variables that occur beneath $\mu$-binders.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus our construction of $\Val$ is terminating.

@@

Given $\Val$, the rules for cyclicity and bisimilarity (shown in Figure~\ref{fig:ab-rules}) can be stated as follows.
For cyclicity, if there is an equivalence class $\ec{t_1}$ of datatype type whose class of values $\Val \ec{t_1} = \mu x. t_2$ are cyclic,
as indicated by the assumption $x \in FV( t_2 )$,
then we can conclude that $\Ec$ is unsatisfiable.
For bisimilarity, if there are two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
of co-datatype type whose classes of values $\Val \ec{t_1}$ and $\Val \ec{t_2}$ are equivalent for a capture-avoiding renaming of $\mu$-bound variables,
then we can conclude $t_1 \teq t_2$.
We demonstrate the bisimilarity rule concretely with the following example.

>>>>

% change?
\newcommand{\thO}{T_{\mathrm{o}}}
\newcommand{\thD}{T_{\mathrm{cd}}}

@@

\section{%The Theory of
(Co)datatypes}
\label{sec:the-theory-of-co-datatypes}

@@

%\paragraph{Characterization.}
One way to define datatypes semantically is as the initial model of the
selector--constructor equations \cite{barrett-et-al-2010}.
\begin{report}
A drawback of this approach is that it does
not naturally account for selectors applied to wrong constructors. Barrett et
al.\ address this by parameterizing the construction by default values, but
this gives rise to spurious equalities between unrelated terms---e.g.,
$\const{s}_{11}(\const{C}_2) \teq \const{s}_{11}(\const{C}_3)$. This flaw
could be corrected, but the added complexity seems to suggest that selectors
are better characterized axiomatically.

@@

Datatypes are additionally characterized by an induction axiom schema\begin{report}:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\,i,j}\, \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\,k}\, \mathit{IH}_{i\negvthinspace j}^k[x_k]\bigr) \longrightarrow P_i[C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\,i}\, P_i[v_i]$}
\DP
\end{aligned}
\]
where the induction hypothesis $\mathit{IH}_{i\negvthinspace j}^k(x)$
denotes either $P_{\iPrime}(x)$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$\end{report}.
%
The axiom schema ensures that the interpretation of datatypes
is a standard model. For example,
for a datatype of natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction prohibits nonstandard models, which could contain cyclic values---e.g.,
an $n$ such that $n \teq \const{S}(n)$---or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.
%
\begin{report}\par\end{report}
%
For codatatypes, the dual notion is called coinduction. \begin{report}It depends on
witnesses $R_i$ that are required to be bisimulations:
%
\[
\begin{aligned}[t]
\text{Coinduction:}\kern.8em %%% TYPESETTING: should be \quad
\AXC{\strut$\begin{gathered}\textstyle \bigwedge_{\,i}\, R_i[v_i, w_i] \\[-\jot]\textstyle
\bigwedge_{\,i}\, \forall v\; w.\;\, R_i[v, w] \longrightarrow
  \bigwedge_{\,j}\, \const{d}_j(v) \teq \const{d}_j(w)
  \mathrel\land
  \const{d}_j(v) \longrightarrow \bigwedge_{\,k}\, \const{s}_{i\negvthinspace j}^{\,k}(v) \sim \const{s}_{i\negvthinspace j}^{\,k}(w)
\end{gathered}$}
\UIC{\strut$\bigwedge_{\,i}\, v_i \teq w_i$}
\DP
\end{aligned}
\]
where $x \sim y$ denotes either $R_{\iPrime}[x, y]$ if there exists some
$\iPrime$ such that the formula is type-correct or $x \teq y$ otherwise.
\end{report}%
This axiom schema guarantees that two values that yield the same
observations must be equal, where the observations are made by using the
selectors and discriminators.
%
\begin{report}\par\end{report}
  %
In addition, codatatypes are guaranteed to contain all values corresponding to
infinite ground constructor terms.
\begin{report}
In general, this cannot be captured by a
first-order axiomatization, since there may be uncountably many of them.
For example, $\ty{stream}_{\,\ty{int}}$ is isomorphic to the uncountable
function space $\ty{nat} \to \ty{int}$.
\end{report}

@@

The decision procedure for the ground theory of (co)datatypes $\thD$ determines the
$\thD$-satisfiability of sets of ground constraints over a fixed
signature~$\Sig$. It is presented as a calculus, whose rules can be applied in
any order until saturation.

The calculus consists of derivation rules that operate on a set $\Ec$ of
equalities and disequalities between $\Sig$-terms, where $\Ec$ is interpreted
as a conjunction.
We write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
are simplified to $t_i$, and moreover assume that all %additional
constraints added to $\Ec$ are normalized in this manner.

To simplify the presentation, we make the following assumptions about $\Sig$.
First, all codatatypes in $\Codata$ are corecursive. This is not restrictive
because noncorecursive codatatypes can be considered as nonrecursive
datatypes.
Second, all types $\tau \in \Nondata$ have infinite cardinality.
This is not overly restrictive, because each interpreted type $\tau \in \Nondata$
having finite cardinality~$n$
can be replaced by an enumeration datatype with $n$~nullary constructors.
Moreover, since the constraints are ground, they cannot entail an upper bound
on the cardinality of any uninterpreted type $\tau \in \Nondata$; thus,
we may without loss of completeness treat all other types as infinite.

@@

A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains an occurrence of a bound variable.
The notation $t \aequiv u$, expressing $\alpha$-\emph{equivalence},
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables.
For example, 
%$\mu x. \const{C}( x ) \aequiv \mu y. \const{C}( y )$ and
$\mu x.\; \const{D}( y, x ) \aequiv \mu z.\; \const{D}( y, z )$,
but
$\mu x.\; \const{C}( x ) \not\aequiv \mu x.\; \const{D}( y, x ) \not\aequiv \mu x.\; \const{D}( z, x )
\not\aequiv \mu y.\; \const{D}( y, x )$.
Two $\mu$-terms may represent the same value while not being $\alpha$-equivalent, as in
$\mu x.\; \const{C}( \const{C}( x ) ) \not\aequiv \mu x.\; \const{C}( x )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

Intuitively, a $\mu$-term $t$ of type $\tau$ describes a class of values of $\tau$, 
%either in the case when $\tau$ is a codatatype or datatype type,
where a cyclic $t$ describes an infeasible class of values when
%$\tau \not\in \Codata$.
$\tau \in \Data$.
At the beginning of this step, 
we compute a mapping $\Val$ from equivalence classes to $\mu$-terms 
that describe the class of values that terms in that equivalence class can take in the models of $\Ec$. 
For each equivalence class $\ec{x}$, we associate a fresh variable $v_{\ec{x}}$ of the same type as $x$ not occurring in $\tEc$,
and initially say that $\Val$ contains $\ec{x} \mapsto v_{\ec{x}}$,
in other words, there are initially no constraints on the values for any equivalence class $\ec{x}$.
Then, $\Val$ is modified by applying the following update exhaustively:
\[
\inferrule{
  v_{\ec{x}} \in FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} ) \}
}
\]
Here, we write $FV( \Val )$ to denote the \emph{free variables} occurring in $\Val$, which
excludes variables that occur beneath $\mu$-binders.
It is straightforward to see that the height of terms produced as a result of this update
is bounded by the number of equivalence classes of $\Ec$,
and thus our construction of $\Val$ is terminating.

@@

Given $\Val$, the rules for cyclicity and bisimilarity (shown in Figure~\ref{fig:ab-rules}) can be stated as follows.
For cyclicity, if there is an equivalence class $\ec{t_1}$ of datatype type whose class of values $\Val \ec{t_1} = \mu x. t_2$ are cyclic,
as indicated by the assumption $x \in FV( t_2 )$,
then we can conclude that $\Ec$ is unsatisfiable.
For bisimilarity, if there are two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
of codatatype type whose classes of values $\Val \ec{t_1}$ and $\Val \ec{t_2}$ are equivalent for a capture-avoiding renaming of $\mu$-bound variables,
then we can conclude $t_1 \teq t_2$.
We demonstrate the bisimilarity rule concretely with the following example.
